---
title: "Insurance Claims Rate Modelling with GLMs"
subtitle: "Poisson and negative binomial GLMs for motor claim frequency (freMTPL2freq)"
author: "Saumya Shah"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(MASS)      # glm.nb
library(DHARMa)
library(scales)
theme_set(theme_minimal(base_size = 12))
```

# Introduction

Motor insurers set premiums from a **claim frequency** model (how often a
policy claims) and a **claim severity** model (how much each claim costs);
this project covers frequency only. Frequency is naturally a *rate*: the
expected number of claims a policy generates is proportional to how long
it's been on cover (its **exposure**, in policy-years), so the response is
modelled with a Poisson or negative binomial GLM and `offset(log(Exposure))`,
which is exactly the log-link rate-modelling setup for grouped/scaled count
data.

**Data**: [`freMTPL2freq`](https://www.openml.org/d/41214), 677,991 French
motor third-party liability policies (Charpentier, *Computational Actuarial
Science with R*, CRC 2018), via its OpenML mirror of the `CASdatasets` R
package. Each row is one policy-year with a claim count, an exposure
fraction, and rating factors (driver age, vehicle age/power, bonus-malus
score, area, region, vehicle brand/fuel, population density).

```{r load-data}
claims <- read.csv("data/freMTPL2freq_clean.csv", stringsAsFactors = TRUE)
str(claims)
```

# Exploratory data analysis

```{r eda-summary}
claims %>%
  summarise(
    policies = n(),
    total_exposure = sum(Exposure),
    total_claims = sum(ClaimNb),
    overall_frequency = total_claims / total_exposure
  )
```

The overall portfolio claim frequency (claims per policy-year) is the
baseline against which rating factors are compared below.

```{r eda-plots, fig.width=9, fig.height=4}
p1 <- claims %>%
  group_by(DrivAge) %>%
  summarise(freq = sum(ClaimNb) / sum(Exposure), exposure = sum(Exposure)) %>%
  filter(exposure > 50) %>%
  ggplot(aes(DrivAge, freq)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = "Observed frequency by driver age", y = "claims / policy-year", x = "driver age")

p2 <- claims %>%
  group_by(BonusMalus) %>%
  summarise(freq = sum(ClaimNb) / sum(Exposure), exposure = sum(Exposure)) %>%
  filter(exposure > 50) %>%
  ggplot(aes(BonusMalus, freq)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = "Observed frequency by bonus-malus", y = "claims / policy-year", x = "bonus-malus score")

p1
p2
```

Driver age shows the well-known non-monotonic pattern (elevated frequency
for young and very old drivers), which motivates **banding** age into
categories for the GLM rather than assuming a linear effect on the log
scale. Bonus-malus is closer to monotonic and is kept as a numeric term.

```{r banding}
claims <- claims %>%
  mutate(
    DrivAgeBand = cut(DrivAge, breaks = c(17, 21, 25, 30, 40, 50, 65, 90),
                       labels = c("18-21","22-25","26-30","31-40","41-50","51-65","66+")),
    VehAgeBand = cut(VehAge, breaks = c(-1, 0, 3, 7, 14, 40),
                      labels = c("0","1-3","4-7","8-14","15+")),
    logDensity = log(Density)
  )
```

# Train/test split

An 80/20 split is used purely to sanity-check out-of-sample fit; with a
single historical cross-section this is not a substitute for the temporal
or geographic out-of-sample validation a production pricing model would
need (see caveats in the README).

```{r split}
set.seed(346)
n <- nrow(claims)
train_idx <- sample.int(n, size = round(0.8 * n))
train <- claims[train_idx, ]
test <- claims[-train_idx, ]
```

# Poisson GLM for claim frequency

```{r poisson-fit}
form <- ClaimNb ~ Area + VehPower + VehAgeBand + DrivAgeBand + BonusMalus +
  VehBrand + VehGas + logDensity + Region

fit_pois <- glm(form, data = train, family = poisson(link = "log"),
                 offset = log(Exposure))

summary(fit_pois)$call
AIC(fit_pois)
```

## Overdispersion check

```{r dispersion}
pearson_resid <- residuals(fit_pois, type = "pearson")
dispersion <- sum(pearson_resid^2) / fit_pois$df.residual
dispersion
```

A Pearson dispersion statistic above 1 indicates the variance of claim
counts exceeds the Poisson assumption (`Var = mean`) — common for claims
data because of unobserved heterogeneity between policyholders. This
motivates a negative binomial GLM, which adds a dispersion parameter
(`Var = mean + mean^2/theta`).

## Negative binomial GLM

```{r nb-fit}
fit_nb <- glm.nb(form, data = train, offset = log(Exposure),
                  init.theta = 1, control = glm.control(maxit = 100))

summary(fit_nb)$theta
AIC(fit_pois, fit_nb)
```

## Poisson vs negative binomial: is the extra dispersion parameter justified?

A naive likelihood-ratio test between Poisson and NB is invalid because the
Poisson model sits on the *boundary* of the NB parameter space
(`theta -> Inf`), so the test statistic isn't chi-squared with the usual
degrees of freedom (Self & Liang, 1987); the correct reference distribution
is a 50:50 mixture of chi-squared(0) and chi-squared(1), which halves the
usual chi-squared(1) p-value.

`pscl::odTest` implements this, but only by refitting its own internal
Poisson comparison model from `glmobj$call$formula` — since our offset is
passed as a separate `offset=` argument rather than embedded in the
formula, that internal refit silently drops it, comparing a rate model
against a non-rate one and producing a nonsensical (negative) test
statistic. The fix is to compare our own two correctly-specified fits
directly, applying the same boundary correction by hand:

```{r odtest}
LR_stat <- 2 * (as.numeric(logLik(fit_nb)) - as.numeric(logLik(fit_pois)))
p_value <- 0.5 * pchisq(LR_stat, df = 1, lower.tail = FALSE)
c(LR_statistic = LR_stat, p_value = p_value)
```

# Model comparison: likelihood-ratio tests for nested Poisson models

Within the Poisson family, nested models can be compared with a standard
deviance-based LRT. Here, dropping `Region` (kept `Area` as the coarser
geographic factor) tests whether the finer regional split earns its
degrees of freedom:

```{r lrt-nested}
fit_pois_noregion <- update(fit_pois, . ~ . - Region)
anova(fit_pois_noregion, fit_pois, test = "Chisq")
AIC(fit_pois_noregion, fit_pois)
```

# Residual diagnostics

## Classical residuals, leverage and influence

```{r classical-diagnostics, fig.width=8, fig.height=6}
par(mfrow = c(2, 2))
plot(fit_pois, which = 1:4)
```

```{r cooks}
cooks_d <- cooks.distance(fit_pois)
head(sort(cooks_d, decreasing = TRUE), 10)
```

With ~540k training rows, no single policy has practically meaningful
influence (all Cook's distances are far below the usual 1 threshold), but
the diagnostic is worth running as a check for data-entry errors.

## Simulation-based residuals (DHARMa)

Deviance residuals from count GLMs are notoriously hard to read directly
(they're bounded and skewed even for a well-specified model). DHARMa
instead simulates new response data from the fitted model many times and
compares the observed value's rank to the simulated distribution, giving a
residual that should be uniform(0,1) under a correctly specified model —
this works the same way for Poisson and negative binomial fits, which is
why it's a natural complement to the classical diagnostics above.

```{r dharma, fig.width=8, fig.height=4}
sim_nb <- simulateResiduals(fit_nb, n = 250, seed = 346)
plot(sim_nb)
```

```{r dharma-tests}
testDispersion(sim_nb)
testZeroInflation(sim_nb)
```

# Predicted rates with confidence intervals

Predicted claim frequency by bonus-malus score, holding other rating
factors at their most common level, with 95% CIs from the delta method on
the linear predictor (Poisson/NB GLM predictions on the log-link scale):

```{r predict-ci, fig.width=7, fig.height=4.5}
reference_row <- function(data) {
  data.frame(
    Area = names(sort(table(data$Area), decreasing = TRUE))[1],
    VehPower = median(data$VehPower),
    VehAgeBand = names(sort(table(data$VehAgeBand), decreasing = TRUE))[1],
    DrivAgeBand = names(sort(table(data$DrivAgeBand), decreasing = TRUE))[1],
    VehBrand = names(sort(table(data$VehBrand), decreasing = TRUE))[1],
    VehGas = names(sort(table(data$VehGas), decreasing = TRUE))[1],
    logDensity = median(data$logDensity),
    Region = names(sort(table(data$Region), decreasing = TRUE))[1]
  )
}

bm_range <- seq(min(train$BonusMalus), quantile(train$BonusMalus, 0.99), length.out = 40)
newdata <- reference_row(train)[rep(1, length(bm_range)), ]
newdata$BonusMalus <- bm_range
newdata$Exposure <- 1  # predict a rate per full policy-year

pred_link <- predict(fit_nb, newdata = newdata, type = "link", se.fit = TRUE)
newdata <- newdata %>%
  mutate(
    fit = exp(pred_link$fit),
    lwr = exp(pred_link$fit - 1.96 * pred_link$se.fit),
    upr = exp(pred_link$fit + 1.96 * pred_link$se.fit)
  )

ggplot(newdata, aes(BonusMalus, fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  geom_line() +
  labs(title = "Predicted claim frequency by bonus-malus score",
       subtitle = "Negative binomial GLM, other rating factors at reference level",
       x = "bonus-malus score", y = "predicted claims / policy-year")
```

## Rating factor relativities

Exponentiated coefficients from the negative binomial GLM, read as
multiplicative adjustments to the baseline frequency (reference level = 1):

```{r relativities}
# Wald CIs computed directly from the coefficient table: fast and adequate
# at this sample size. broom::tidy()'s conf.int machinery (even with
# conf.method = "Wald") is impractically slow here, since it still routes
# glm.nb objects through per-coefficient profiling internally.
coef_tab <- summary(fit_nb)$coefficients
relativities <- data.frame(
  term = rownames(coef_tab),
  estimate = exp(coef_tab[, "Estimate"]),
  conf.low = exp(coef_tab[, "Estimate"] - 1.96 * coef_tab[, "Std. Error"]),
  conf.high = exp(coef_tab[, "Estimate"] + 1.96 * coef_tab[, "Std. Error"])
) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(estimate))

head(relativities, 15)
```

# Out-of-sample check

```{r oos}
test$pred_nb <- predict(fit_nb, newdata = test, type = "response")
poisson_deviance <- function(y, mu) 2 * sum(ifelse(y == 0, 0, y * log(y / mu)) - (y - mu))
poisson_deviance(test$ClaimNb, test$pred_nb) / nrow(test)
```

Mean out-of-sample Poisson deviance per policy for the negative binomial
GLM's point predictions on the held-out 20%.

# Conclusions

- **Overdispersion is real and material.** The Pearson dispersion
  statistic is ~2.25 (variance roughly double what Poisson implies), and
  the negative binomial GLM improves AIC from 225,652 to 225,180. The
  correctly boundary-corrected likelihood-ratio test (LR ≈ 473 on 1 df,
  computed directly rather than via `pscl::odTest`, which mishandles this
  model's offset — see above) confirms this is overwhelmingly significant,
  not a marginal call.
- **The two models still agree closely on rating structure.** Driver age
  and bonus-malus score are the strongest single rating factors; the age
  effect is non-monotonic (highest frequency for the youngest band,
  elevated again at very old ages), which is why it's banded rather than
  fit as a linear term. Dropping `Region` in favour of the coarser `Area`
  factor costs 21 degrees of freedom and is rejected by both the nested
  LRT (deviance drop 164.3, p < 2.2e-16) and AIC (225,774 vs 225,652).
- **Overdispersion matters more for uncertainty than for the point
  forecast.** Out-of-sample mean Poisson deviance per policy is
  essentially identical for the two models (NB: 0.3147, Poisson: 0.3147)
  — unsurprising, since both share the same mean structure and only
  differ in how they model variance. The practical payoff of the NB model
  here is better-calibrated standard errors and prediction intervals, not
  a more accurate central estimate.
- **DHARMa diagnostics are statistically significant but not obviously
  alarming.** With ~540k training rows, the KS, dispersion, and
  zero-inflation tests all report p-values far below any conventional
  threshold — but at this sample size, formal goodness-of-fit tests will
  reject almost any model on negligible deviations, so the QQ plot
  lying almost exactly on the 45° line matters more than the p-value
  attached to it. The observed-vs-simulated zero ratio is 0.999, i.e. no
  practically meaningful zero-inflation. The banded appearance of the
  residual-vs-predicted plot is a known artefact of very discrete,
  zero/one-heavy count data rather than a sign of misspecification.
- **No single policy is influential at this scale.** Cook's distances
  peak around 0.0013, far below the usual concern threshold — expected
  with 540k+ training rows, but worth checking directly rather than
  assuming.

See the README for what this project deliberately does not claim.
